Appendix

### 2025-10-29 manuscript_rmarkdown/manuscript.Rmd
Updates to text.

We used simulated data to assess invariance to location and scale of the proposed power transformations,
to develop the empirical central normality test, and to determine weighting for robust transformations.
The $\lambda$ parameter for conventional power transformations (Eqn. \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}),
as well as $\lambda$, $x_0$ and $s$ parameters for location- and scale-invariant power transformations 
(Eqn. \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA algorithm 
for derivative-free bound constraint optimisation [@Powell2009-zb] through maximum likelihood estimation.
The required algorithms were implemented in the `power.transform` R software package [@Zwanenburg2024-kq] (version 1.0.1).
Of note, the `power.transform` package shifts feature values into the positive domain if negative or zero values are present for Box-Cox power transformations.

@@

Each sequence was created as follows: first $n$ values are sampled from $\mathcal{N}(0, 1)$.
Then, the elements outside the central portion of the sequence (i.e. $\mathbf{X} \setminus \mathbf{X}_{\text{c}}$) are replaced.
Elements with $pi < \frac{1-\kappa}{2}$ are replaced with values sampled from $\mathcal{U}(-10,\min(\mathbf{X}_{\text{c}}))$,
and elements with $p_i > \frac{1 + \kappa}{2}$ are replaced with values sampled from $\mathcal{U}(\max(\mathbf{X}_{\text{c}}), 10)$.

@@

```{r empirical-central-normality-error_rate, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Observed type I error rates for central normality tests in the presence of data sampled from ",
  "central normal distributions with $\\kappa \\in \\left\\{0.6, 0.8, 1.0\\right\\}$. ",
  "Three empirical central normality (ECN) tests were performed, parameterised with same value for the central portion $\\kappa$. ",
  "Additionally, the Shapiro-Wilk test was assessed. ",
  "ECN: empirical central normality"
)

@@

```{r empirical-central-normality-examples, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Bi-modal distributions and empirical central normality test results. ",
  "The feature (black) is a mixture of two identical sample sets (blue and orange, $n = 100$) ",
  "drawn from normal distributions that are offset by a distance $d$. ",
  "The empirical central normality ($\\kappa = 0.80$) is used to test the hypothesis that the mixture is sampled from ",
  "a central normal distribution. As may be observed, with increasing offset $d$ ",
  "the test's p-value decreases. ",
  "Quantile-quantile plots are drawn below each distribution."
)

@@

A common feature in health-related datasets is age.
Here we use data on 228 patients with lung cancer that was collected and published by Loprinzi et al. [@Loprinzi1994-cd].
The age in the cohort was $62.4 \pm 9.1$ (mean ± standard deviation) years.
Applying conventional and invariant Yeo-Johnson transformations to patient age yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation ($\sum r_i = 16.5$, $p = 0.69$);
conventional transformation ($\lambda = 2.0$, $\sum r_i = 11.5$, $\mu_{YJ} = 1.8 \cdot 10^3$, $\sigma_{YJ} = 0.5 \cdot 10^3$, $p = 0.96$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 2.0$, $\sum r_i = 11.5$, $\mu_{YJ} = 1.8 \cdot 10^3$, $\sigma_{YJ} = 0.5 \cdot 10^3$, $p = 0.96$);
location- and scale-invariant transformation ($\lambda = 1.3$, $\sum r_i = 8.8$, $\mu_{YJ} = -1.2$, $\sigma_{YJ} = 1.1$, $p = 0.98$);
and robust location- and scale-invariant transformation ($\lambda = 1.3$, $\sum r_i = 9.3$, $\mu_{YJ} = -1.0$, $\sigma_{YJ} = 1.1$, $p = 0.93$).

Location- and scale-invariant transformation led to a lower overall residual error, indicating a better transformation to normality.
Conventional transformations inflated the mean $\mu_{YJ}$ and standard deviation $\sigma_{YJ}$ of the age feature after transformation.
The empirical central normality test did not detect any statistically significant deviations from central normality for any transformation (all $p \geq 0.93$).

@@

Gorman, Williams and Fraser recorded body mass (in grams) of 342 penguins of three different species [@Gorman2014-eo].
The body mass was $(4.2 \pm 0.8) \cdot 10^3$ (mean ± standard deviation) grams, and not centrally normal ($p < 0.001$).
Applying conventional and invariant Yeo-Johnson transformations to body mass yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation (residual sum $\sum r_i = 48.0$, $p < 0.001$);
conventional transformation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{YJ} = 2.1$, $\sigma_{YJ} = 4 \cdot 10^{-3}$, $p = 0.10$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{YJ} = 2.1$, $\sigma_{YJ} = 4 \cdot 10^{-3}$, $p = 0.10$);
location- and scale-invariant transformation ($\lambda = 0.5$, $\sum r_i = 26.8$, $\mu_{YJ} = 0.9$, $\sigma_{YJ} = 0.9$, $p = 0.28$);
and robust location- and scale-invariant transformation ($\lambda = 0.3$, $\sum r_i = 22.0$, $\mu_{YJ} = 0.7$, $\sigma_{YJ} = 0.9$, $p = 0.69$).

Location- and scale-invariant transformation produced a lower overall residual errors, indicating a better transformation.
Moreover, conventional transformations led to low standard deviation $\sigma_{YJ}$ of the body mass feature after transformation.
The empirical central normality test did not detect any statistically significant deviations from central normality for any transformation (all $p \geq 0.10$).

@@

Every transformation reduced the residual sum.
None of the transformations yielded a centrally normal distribution.
Conventional transformations had high values for the $\lambda$ parameter, which could lead to numerical issues.

@@

The Top Gear dataset contains data on 297 vehicles that appeared on the BBC television show *Top Gear* [@Alfons2021-kc].
Within this dataset, the fuel consumption feature contains outliers due to highly fuel-efficient vehicles.
Applying conventional and invariant Yeo-Johnson transformations to the fuel consumption feature yielded the following results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum $\sum r_i = 54$, $p=0.72$);
conventional transformation ($\lambda = -0.1$, $\sum r_i = 55$, $\mu_{YJ} = 3.0$, $\sigma_{YJ} = 0.3$, $p < 0.001$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 0.8$, $\sum r_i = 48$, $\mu_{YJ} = 29$, $\sigma_{YJ} = 15$, $p=0.37$);
location- and scale-invariant transformation ($\lambda = -1.3$, $\sum r_i = 44$, $\mu_{YJ} = 0.5$, $\sigma_{YJ} = 0.1$, $p < 0.001$);
and robust location- and scale-invariant transformation ($\lambda = 1.0$, $\sum r_i = 56$, $\mu_{YJ} = 1.7$, $\sigma_{YJ} = 1.0$, $p=0.76$).

Outliers cause non-robust transformations to fail to transform the data to a centrally normal distribution
(empirical central normality test $p < 0.001$ for conventional and invariant transformations).
Robust transformations produce distributions that are centrally normal (empirical central normality test $p > 0.05$).

@@

The ischemic stroke dataset contains historic data from 126 patients with risk at ischemic stroke [@Kuhn2019-kt].
These patients underwent Computed Tomography Angiography to characterize the carotid artery blockages.
Angiography imaging was then assessed, and various characteristics related to the blood vessels and the disease are measured.
The maximum arterial wall thickness feature contains several instances with outlier values.
Applying conventional and invariant Yeo-Johnson transformations to this feature yielded the following results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum $\sum r_i = 110$, $p=0.83$);
conventional transformation ($\lambda = -0.7$, $\sum r_i = 30$, $\mu_{YJ} = 1.0$, $\sigma_{YJ} = 0.1$, $p=0.003$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 1.1$, $\sum r_i = 136$, $\mu_{YJ} = 7.2$, $\sigma_{YJ} = 14.3$, $p=0.88$);
location- and scale-invariant transformation ($\lambda = 0.2$, $\sum r_i = 12$, $\mu_{YJ} = -11.8$, $\sigma_{YJ} = 6.9$, $p=0.15$);
and robust location- and scale-invariant transformation ($\lambda = -0.3$, $\sum r_i = 30$, $\mu_{YJ} = 0.8$, $\sigma_{YJ} = 0.2$, $p=0.18$).

The conventional non-robust transformation failed to produce a centrally normal distribution (empirical central normality test $p=0.003$).
Robust transformations produce distributions that are centrally normal (empirical central normality test $p > 0.05$).

>>>>

We used simulated data to assess invariance to location and scale of the proposed power transformations,
to develop the empirical central normality test, and to determine weighting for robust transformations.
The $\lambda$ parameter for conventional power transformations (Eqn. \ref{eqn:box-cox-original} and \ref{eqn:yeo-johnson-original}),
as well as $\lambda$, $x_0$ and $s$ parameters for location- and scale-invariant power transformations 
(Eqn. \ref{eqn:box-cox-invariant} and \ref{eqn:yeo-johnson-invariant}) were estimated using the BOBYQA algorithm 
for derivative-free bound constraint optimisation [@Powell2009-zb] through maximum likelihood estimation.
The required algorithms were implemented in the `power.transform` R software package [@Zwanenburg2024-kq] (version 1.0.2).
Of note, the `power.transform` package shifts feature values into the positive domain if negative or zero values are present for Box-Cox power transformations.

@@

Each sequence was created as follows: first $n$ values are sampled from $\mathcal{N}(0, 1)$.
Then, the elements outside the central portion of the sequence (i.e. $\mathbf{X} \setminus \mathbf{X}_{\text{c}}$) are replaced.
Elements with $p_i < \frac{1-\kappa}{2}$ are replaced with values sampled from $\mathcal{U}(-10,\min(\mathbf{X}_{\text{c}}))$,
and elements with $p_i > \frac{1 + \kappa}{2}$ are replaced with values sampled from $\mathcal{U}(\max(\mathbf{X}_{\text{c}}), 10)$.

@@

```{r empirical-central-normality-error_rate, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Observed type I error rates for central normality tests in the presence of data sampled from ",
  "central normal distributions with $\\kappa \\in \\left\\{0.6, 0.8, 1.0\\right\\}$. ",
  "Three empirical central normality (ECN) tests were performed, parameterised with same value for the central portion $\\kappa$. ",
  "Additionally, the Shapiro-Wilk test was assessed. ",
  "The dashed line indicates the expected type I error rate for rejecting the null hypothesis ",
  "at significance level $\\alpha = 0.05$. ",
  "ECN: empirical central normality"
)

@@

```{r empirical-central-normality-examples, warning=FALSE, message=FALSE, fig.cap=cap, fig.height=2.5}
cap <- paste0(
  "Bi-modal distributions and empirical central normality test results. ",
  "The feature (black) is a mixture of two identical sample sets (blue and orange, $n = 100$) ",
  "drawn from normal distributions that are offset by a distance $d$. ",
  "The empirical central normality test ($\\kappa = 0.80$) is used to test the hypothesis that the mixture is sampled from ",
  "a central normal distribution. As may be observed, with increasing offset $d$ ",
  "the test's p-value decreases. ",
  "Quantile-quantile plots are drawn below each distribution."
)

@@

A common feature in health-related datasets is age.
Here we use data on 228 patients with lung cancer that was collected and published by Loprinzi et al. [@Loprinzi1994-cd].
The age in the cohort was $62.4 \pm 9.1$ (mean ± standard deviation) years.
Applying conventional and invariant Yeo-Johnson transformations to patient age yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation ($\sum r_i = 16.5$, $p = 0.54$);
conventional transformation ($\lambda = 2.0$, $\sum r_i = 11.5$, $\mu_{YJ} = 1.8 \cdot 10^3$, $\sigma_{YJ} = 0.5 \cdot 10^3$, $p = 0.92$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 2.0$, $\sum r_i = 11.5$, $\mu_{YJ} = 1.8 \cdot 10^3$, $\sigma_{YJ} = 0.5 \cdot 10^3$, $p = 0.92$);
location- and scale-invariant transformation ($\lambda = 1.3$, $\sum r_i = 8.8$, $\mu_{YJ} = -1.2$, $\sigma_{YJ} = 1.1$, $p = 0.95$);
and robust location- and scale-invariant transformation ($\lambda = 1.3$, $\sum r_i = 9.3$, $\mu_{YJ} = -1.0$, $\sigma_{YJ} = 1.1$, $p = 0.87$).

Location- and scale-invariant transformation led to a lower overall residual error, indicating a better transformation to normality.
Conventional transformations inflated the mean $\mu_{YJ}$ and standard deviation $\sigma_{YJ}$ of the age feature after transformation.
The empirical central normality test did not detect any statistically significant deviations from central normality for any transformation (all $p \geq 0.87$).

@@

Gorman, Williams and Fraser recorded body mass (in grams) of 342 penguins of three different species [@Gorman2014-eo].
The body mass was $(4.2 \pm 0.8) \cdot 10^3$ (mean ± standard deviation) grams, and not centrally normal ($p < 0.001$).
Applying conventional and invariant Yeo-Johnson transformations to body mass yielded the following results, see Figure \ref{fig:experimental-results-invariance}:
no transformation (residual sum $\sum r_i = 48.0$, $p < 0.001$);
conventional transformation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{YJ} = 2.1$, $\sigma_{YJ} = 4 \cdot 10^{-3}$, $p = 0.03$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = -0.5$, $\sum r_i = 32.2$, $\mu_{YJ} = 2.1$, $\sigma_{YJ} = 4 \cdot 10^{-3}$, $p = 0.03$);
location- and scale-invariant transformation ($\lambda = 0.5$, $\sum r_i = 26.8$, $\mu_{YJ} = 0.9$, $\sigma_{YJ} = 0.9$, $p = 0.12$);
and robust location- and scale-invariant transformation ($\lambda = 0.3$, $\sum r_i = 22.0$, $\mu_{YJ} = 0.7$, $\sigma_{YJ} = 0.9$, $p = 0.48$).

Location- and scale-invariant transformation yielded lower overall residual errors, indicating a better transformation to central normality,
which is reflected by the empirical central normality test results. 
Conventional transformations led to low standard deviation $\sigma_{YJ}$ of the body mass feature after transformation,
and the null hypothesis of the empirical central normality test was rejected (at significance level $\alpha = 0.05$) in both cases.

@@

Every transformation reduced the residual sum compared to the original data.
None of the transformations yielded a centrally normal distribution.
Conventional transformations had high values for the $\lambda$ parameter, which could lead to numerical issues.

@@

The Top Gear dataset contains data on 297 vehicles that appeared on the BBC television show *Top Gear* [@Alfons2021-kc].
Within this dataset, the fuel consumption feature contains outliers due to highly fuel-efficient vehicles.
Applying conventional and invariant Yeo-Johnson transformations to the fuel consumption feature yielded the following results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum $\sum r_i = 54$, $p=0.55$);
conventional transformation ($\lambda = -0.1$, $\sum r_i = 55$, $\mu_{YJ} = 3.0$, $\sigma_{YJ} = 0.3$, $p < 0.001$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 0.8$, $\sum r_i = 48$, $\mu_{YJ} = 29$, $\sigma_{YJ} = 15$, $p=0.21$);
location- and scale-invariant transformation ($\lambda = -1.3$, $\sum r_i = 44$, $\mu_{YJ} = 0.5$, $\sigma_{YJ} = 0.1$, $p < 0.001$);
and robust location- and scale-invariant transformation ($\lambda = 1.0$, $\sum r_i = 56$, $\mu_{YJ} = 1.7$, $\sigma_{YJ} = 1.0$, $p=0.59$).

Outliers caused non-robust transformations to fail the empirical central normality test for
transforming the feature to a centrally normal distribution.

@@

The ischemic stroke dataset contains historic data from 126 patients with risk at ischemic stroke [@Kuhn2019-kt].
These patients underwent Computed Tomography Angiography to characterize the carotid artery blockages.
Angiography imaging was then assessed, and various characteristics related to the blood vessels and the disease are measured.
The maximum arterial wall thickness feature contains several instances with outlier values.
Applying conventional and invariant Yeo-Johnson transformations to this feature yielded the following results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum $\sum r_i = 110$, $p=0.73$);
conventional transformation ($\lambda = -0.7$, $\sum r_i = 30$, $\mu_{YJ} = 1.0$, $\sigma_{YJ} = 0.1$, $p < 0.001$);
Raymaekers and Rousseeuw's robust adaptation ($\lambda = 1.1$, $\sum r_i = 136$, $\mu_{YJ} = 7.2$, $\sigma_{YJ} = 14.3$, $p=0.80$);
location- and scale-invariant transformation ($\lambda = 0.2$, $\sum r_i = 12$, $\mu_{YJ} = -11.8$, $\sigma_{YJ} = 6.9$, $p=0.07$);
and robust location- and scale-invariant transformation ($\lambda = -0.3$, $\sum r_i = 30$, $\mu_{YJ} = 0.8$, $\sigma_{YJ} = 0.2$, $p=0.09$).

For conventional non-robust transformation the null hypothesis was rejected (at $\alpha = 0.05$), indicating
that they failed to produce a transformed feature that followed a central normal distribution.

### 2025-10-29 manuscript_rmarkdown/manuscript_appendix.Rmd
Updates to text.

Table \ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual} shows
the average normalised residual error for these transformation methods. This
metric is computed from the residual error and the residual error of the central
portion ($\kappa = 0.70$) for sequences without and with outliers, respectively.
In each experiment, the respective residual error was normalised by dividing with
the sequence length, and can be interpreted as the average residual error of a
single sample. This reduces the effect of sequence length on the residual error.

@@

\midrule
none                                  & 16.5 & 110.1 & 54.5 & 328.4 & 48.0 \\
conventional                          & 11.5 &  30.0 & 55.3 & 319.0 & 32.2 \\
conventional (z-score norm.)          & 11.5 &  19.3 & 49.0 & 326.2 & 33.3 \\
conventional (robust scaling)         & 11.3 &  33.3 & 55.9 & 326.4 & 31.5 \\
Raymaekers-Rousseeuw                  & 11.5 & 136.7 & 47.7 & 315.1 & 32.2 \\
Raymaekers-Rousseeuw (z-score norm.)  & 13.2 & 214.5 & 57.4 & 326.2 & 33.3 \\
Raymaekers-Rousseeuw (robust scaling) & 13.4 & 203.0 & 56.9 & 326.4 & 31.5 \\
invariant                             &  8.8 &  12.2 & 44.0 & 326.4 & 26.8 \\
robust invariant                      &  9.3 &  30.2 & 55.8 & 308.1 & 22.0 \\

>>>>

Table \ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual} shows
the average normalised residual error for these transformation methods. This
metric is computed from the residual error and the residual error of the central
portion ($\kappa = 0.80$) for sequences without and with outliers, respectively.
In each experiment, the respective residual error was normalised by dividing with
the sequence length, and can be interpreted as the average residual error of a
single sample. This reduces the effect of sequence length on the residual error.

@@

\midrule
none                                  & 16.5 & 110.1 & 54.5 & 328.4 & 48.0 \\
conventional                          & 11.5 &  30.0 & 55.3 & 319.0 & 32.2 \\
conventional (z-score norm.)          & 11.5 &  19.3 & 49.0 & 326.2 & 33.3 \\
conventional (robust scaling)         & 11.3 &  33.3 & 55.9 & 326.5 & 31.5 \\
Raymaekers-Rousseeuw                  & 11.5 & 136.7 & 47.7 & 315.1 & 32.2 \\
Raymaekers-Rousseeuw (z-score norm.)  & 13.2 & 214.5 & 57.4 & 326.2 & 33.3 \\
Raymaekers-Rousseeuw (robust scaling) & 13.4 & 202.9 & 56.9 & 326.4 & 31.5 \\
invariant                             &  8.8 &  12.2 & 44.0 & 326.4 & 26.8 \\
robust invariant                      &  9.3 &  30.2 & 55.8 & 308.1 & 22.0 \\
